---
title: "Covid SIR models project"
author: "Oliver Laufenberg"
date: "2023-11-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This project uses Wisconsin DHS (Department of Health Services) data to compare different SIR model variants

This first section reads in the data from the provided hospitalization data between April 1st 2021, and December 31st 2022
```{r}
rm(list = ls()) #clear all
library(readxl)
Covid_data = read_excel("Covid HX ICU Patients Table.xlsx")
Covid_data = head(Covid_data, 728) #first two years only
Covid_data = tail(Covid_data, (nrow(Covid_data) - 88)) #Starting in april, 88 days after the start of the year, the data was weird before then
Covid_data

Hospital_data = Covid_data[5] #Using only the inpatient hospitalizations
plot(1:nrow(Hospital_data), t(Hospital_data))

```


```{r}
# Parameters
beta=0.027   # infection rate
gamma=0.007  # recovery rate
I0=(Hospital_data[[1]][1])/N    # initial proportion of infected individuals
S0=1-I0    # initial proportion of susceptible individuals
R0=0       # initial proportion of recovered individuals
N=5896      # total population of Wisconsin, 5.896 million
days=nrow(Hospital_data)    # simulation time in days

# Initialize vector for S, I, and R
SIR=matrix(0, ncol = 3, nrow = days)
SIR[1, ]=c(S0 * N, I0 * N, R0 * N)

# Run SIR model
for (t in 2:days) {
  dS=-beta * SIR[t-1, 1] * SIR[t-1, 2] / N
  dI=(beta * SIR[t-1, 1] * SIR[t-1, 2] / N) - gamma * SIR[t-1, 2]
  dR=gamma * SIR[t-1, 2]
  
  SIR[t, ]=SIR[t-1, ] + c(dS, dI, dR)
}

# Plotting the results
library(ggplot2)

sir_result=data.frame(Time = 1:days, SIR)
sir_result = cbind(sir_result,Hospital_data) #attaching hospitalization data to plot against model
colnames(sir_result)[2:5]=c("Susceptible", "Infected", "Recovered", "Hospitalizations")
head(sir_result)

ggplot(sir_result, aes(x = Time)) +
  geom_point(aes(y = Susceptible, color = "Susceptible", shape = "Susceptible")) +
  geom_point(aes(y = Infected, color = "Infected", shape = "Infected")) +
  geom_point(aes(y = Recovered, color = "Recovered", shape = "Recovered")) +
  geom_point(aes(y = Hospitalizations, color = "Hospitalizations", shape = "Hospitalizations")) +
  labs(title = "SIR Model",
       x = "Time",
       y = "People in each group, 100 thousands (excluding hospitalizations)")
  #scale_color_manual(values = c("Susceptible" = "blue", "Infected" = "red", "Recovered" = "green"))


```
Here this demonstrates that an SIR model doesn't really capture the full picture, but can be decent if we choose the right coefficients and scale

To add some additional complexity, we can examine an SIER model, where individuals are exposed to a disease, but not infectious to others for a time period
```{r}
# Parameters
beta=0.027   # infection (exposure) rate
gamma=0.007  # recovery rate
sigma=0.3   # exposed becoming infectious rate
I0=(Hospital_data[[1]][1])/N    # initial proportion of infected individuals
S0=1-I0    # initial proportion of susceptible individuals
R0=0       # initial proportion of recovered individuals
E0=0       # initial proportion of exposed indivuduals
N=5896      # total population of Wisconsin, 5.896 million
days=nrow(Hospital_data)    # simulation time in days

# Initialize vector for S, I, and R
SIR=matrix(0, ncol = 4, nrow = days)
SIR[1, ]=c(S0 * N, E0 * N, I0 * N, R0 * N)

# Run SIR model
for (t in 2:days) {
  dS=-beta * SIR[t-1, 1] * SIR[t-1, 3] / N
  dE=(beta * SIR[t-1, 1] * SIR[t-1, 3] / N) - (sigma * SIR[t-1, 2])
  dI=(sigma * SIR[t-1, 2]) - (gamma * SIR[t-1, 3])
  dR=gamma * SIR[t-1, 3]
  
  SIR[t, ]=SIR[t-1, ] + c(dS, dE, dI, dR)
}


# Plotting the results
library(ggplot2)

sir_result=data.frame(Time = 1:days, SIR)
sir_result = cbind(sir_result,Hospital_data) #attaching hospitalization data to plot against model
colnames(sir_result)[2:6]=c("Susceptible", "Exposed", "Infected", "Recovered", "Hospitalizations")
head(sir_result)

ggplot(sir_result, aes(x = Time)) +
  geom_point(aes(y = Susceptible, color = "Susceptible", shape = "Susceptible")) +
  geom_point(aes(y = Exposed, color = "Exposed", shape = "Exposed")) + 
  geom_point(aes(y = Infected, color = "Infected", shape = "Infected")) +
  geom_point(aes(y = Recovered, color = "Recovered", shape = "Recovered")) +
  geom_point(aes(y = Hospitalizations, color = "Hospitalizations", shape = "Hospitalizations")) +
  labs(title = "SEIR Model",
       x = "Time",
       y = "People in each group, 100 thousands (excluding hospitalizations)")


```
Not really very different, 


Next we're going to look at an SIRI model, where people can be infected again after recovering
```{r}
# Parameters
beta=0.027   # infection rate
gamma=0.007  # recovery rate
betaPrime = 0.0027 #infection rate of people who have been infected previously
I0=(Hospital_data[[1]][1])/N    # initial proportion of infected individuals
S0=0.99    # initial proportion of susceptible individuals
R0=0       # initial proportion of recovered individuals
Iprime0=0  # initial proportion of individuals infected more than once
Rprime0=0  # initial population of individuals recovered more than once
N=5896      # total population of Wisconsin, 5.896 million
days=nrow(Hospital_data)    # simulation time in days

# Initialize vector for S, I, and R
SIRI=matrix(0, ncol = 4, nrow = days)
SIRI[1, ]=c(S0 * N, I0 * N, R0 * N, Iprime0 * N)

# Run SIR model
for (t in 2:days) {
  dS=-((beta * SIRI[t-1, 1]) * (SIRI[t-1, 2] + SIRI[t-1, 4])/ N)
  dI=((beta * SIRI[t-1, 1]) * (SIRI[t-1, 2] + SIRI[t-1, 4])/ N) - gamma * SIRI[t-1, 2]
  dR=(gamma * SIRI[t-1, 2]) + (gamma * SIRI[t-1, 4]) - ((betaPrime * SIRI[t-1, 3]) * (SIRI[t-1, 2] + SIRI[t-1, 4]) / N)
  dIprime=((betaPrime * SIRI[t-1, 3]) * (SIRI[t-1, 2] + SIRI[t-1, 4]) / N) - gamma * SIRI[t-1, 4]
  SIRI[t, ]=SIRI[t-1, ] + c(dS, dI, dR, dIprime)
}

# Plotting the results
library(ggplot2)

sir_result=data.frame(Time = 1:days, SIRI)
sir_result = cbind(sir_result,Hospital_data) #attaching hospitalization data to plot against model
colnames(sir_result)[2:6]=c("Susceptible", "Infected", "Recovered", "Reinfected", "Hospitalizations")
sir_result$Infected = sir_result$Reinfected + sir_result$Infected
head(sir_result)
tail(sir_result)

ggplot(sir_result, aes(x = Time)) +
  geom_point(aes(y = Susceptible, color = "Susceptible", shape = "Susceptible")) +
  geom_point(aes(y = Infected, color = "Infected", shape = "Infected")) +
  geom_point(aes(y = Recovered, color = "Recovered", shape = "Recovered")) +
  geom_point(aes(y = Hospitalizations, color = "Hospitalizations", shape = "Hospitalizations")) +
  #geom_point(aes(y = MultiInfected, color = "MultiInfected", shape = "MultiInfected")) +
  labs(title = "SIRI Model",
       x = "Time",
       y = "People in each group, 100 thousands (excluding hospitalizations)")


```

